---
title: "Wholesale Consumer Segmentation with R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Business Problem

Company XYZ is a wholesale distributor that serves a number of business clients(e.g., retailers, restaurant,
hotels, or cafes) in Portugal.The management of XYZ would like to get a deeper understanding of the spending
patterns of their clients, so that they can make strategic decisions to better meet the clients’ demand.

#### 1.1 Business Task:
  • Identify meaningful spending patterns of Company XYZ’s wholesale clients and come up with strategies
that better meet their client’s demand.

### 2. Prepare Data
#### 2.1 About the data

• Wholesale customers dataset from Company XYZ
• The dataset consists of 440 clients and their annual spending
• Each record corresponds to a client, and contains a client’s region, channel, and annual spending across 6 product categories:
      • Channel
          1. Horeca (Hotel/Restaurant/Cafe)
          2. Retail
      • Region
          1. Lisbon
          2. Oporto
          3. other regions
      • Product categories
          1. Fresh
          2. Milk
          3. Grocery
          4. Frozen
          5. Detergents_Paper
          6. Delicatessen

### 3. Process Data

#### 3.1 Import data and required packages
```{r loading_packages, message = FALSE}
library(dplyr)
library(arules)
library(stats)
library(cluster)
library(ggplot2)
library(factoextra)
library(skimr)
library(tidyr)
```


```{r , comment= TRUE}
# import the dataset
wholesale = read.csv("Wholesale customers data.csv")

# Get summary of data to look at quartile
summary(wholesale)

```
* From the summary, we get the Q1, Median, and Q3 for each product categories.

#### 3.2 Data Exploration

```{r }
channel_count=wholesale %>% 
  mutate(channel_name = case_when(Channel==1 ~ 'Hotel_Restaurant_Cafe',
                                  Channel==2 ~ 'Retail')) %>%
  group_by(channel_name) %>% summarise(n=n())

# Plotting clients count vs channels
ggplot(channel_count, aes(y=n,x=channel_name,fill=channel_name))+
  geom_bar(stat='identity' ) +
  scale_fill_manual(values = c('darkblue','cyan'))+
  labs(x='Channel', y='total_clients')+
  geom_text(aes(label=n),vjust=0)
```
```{r }
# Plotting clients count vs region
region_count=wholesale %>% 
  mutate(region_name = case_when(Region==1 ~ 'Lisbon',Region==2 ~ 'Oporto',
                                 Region==3 ~ 'Other')) %>%
           group_by(region_name) %>% summarise(n=n())

ggplot(region_count, aes(y=n,x=as.character(region_name),fill=factor(region_name)))+
  geom_bar(stat='identity' ) +
  scale_fill_manual(values = c('burlywood2','coral2','darkcyan'))+
  labs(x='Region', y='total_clients')+
  geom_text(aes(label=n),vjust=0)

```  
Overall, we found that the majority of the clients are from the 'Other' region 
and most of the clients are from the Horeca channel. To further understand the
relationships within the data and the grouping of the clients for focused 
targeting, association rules and clustering analysis have been performed.


### 4. Analyzing Data

#### 4.1 Association Rule

This is a technique that helps uncover interesting co-occurrence and co-purchase 
habits. We are mainly focusing on the high spending clients (top 25% annual 
spending for each product category), since they are the clients who bring in the
majority of the revenue (62%).

```{r}
# 1. Utilizing association rules to identify if there's any meaningful 
#relationship between:
#- Annual spending of each product categories
#- Channel and spending of each product
#- between products

#To do that, we first need to transform them into binary variables:
#  - round(Min - Q1): considered as low spending
# - round(Q1 - Q3): considered as medium spending
# - round(Q3 - Max): considered as high spending

# Convert all variables to binary variable in order to use Arules package. 
wholesale_convert = wholesale %>%
  mutate(RLisbon = ifelse(Region == 1, 1, 0),
         ROporto = ifelse(Region == 2, 1, 0),
         ROther = ifelse(Region == 3, 1, 0)) %>%
  
  mutate(CHoreca = ifelse(Channel == 1, 1, 0),
         CRetail = ifelse(Channel == 2, 1, 0)) %>%
  
  mutate(FreshH = ifelse(Fresh > 17000, 1, 0),
         FreshM = ifelse(Fresh <= 17000 & Fresh > 3100, 1, 0),
         FreshL = ifelse(Fresh <= 3100, 1, 0)) %>%
  
  mutate(MilkH = ifelse(Milk > 7200, 1, 0),
         MilkM = ifelse(Milk <= 7200 & Milk > 1500, 1, 0),
         MilkL = ifelse(Milk <= 1500, 1, 0)) %>%
  
  mutate(GroceryH = ifelse(Grocery > 11000, 1, 0) ,
         GroceryM = ifelse(Grocery <= 11000 & Grocery > 2200, 1, 0),
         GroceryL = ifelse(Grocery <= 2200, 1, 0)) %>%
  
  mutate(FrozenH = ifelse(Frozen > 4000, 1, 0),
         FrozenM = ifelse(Frozen <= 4000 & Frozen > 800, 1, 0),
         FrozenL = ifelse(Frozen <= 800, 1, 0)) %>%
  
  mutate(DetergentsH = ifelse(Detergents_Paper > 4000, 1, 0),
         DetergentsM = ifelse(Detergents_Paper <= 4000 & Detergents_Paper > 300,
                              1, 0),
         DetergentsL = ifelse(Detergents_Paper <= 300, 1, 0)) %>%
  
  mutate(DelicatessenH = ifelse(Delicatessen > 1900, 1, 0),
         DelicatessenM = ifelse(Delicatessen <= 1900 & Delicatessen > 410, 1, 0),
         DelicatessenL = ifelse(Delicatessen <= 410, 1, 0))

whs_convertion = wholesale_convert[,9:31]
```


To obtain a meaningful amount of association rules, we need to set a minimum 
support and confidence.For this data set, we set a minimum support of 20% and 
a minimum confidence of 70%. If no rules are obtained, support%
will be reduced to see if any meaningful rules can be found.

- support percentage:fraction of transactions containing both attributes within
the data set

- confidence: the ratio of transactions having attribute A that also contains 
the attribute B

```{r }
# When minsupport at 20% and minconfidence at 70%:
whs = whs_convertion %>% mutate(across(,as.logical))
whs_rules = apriori(whs, 
                    parameter = list(supp=0.2,conf=0.7, minlen = 2))

inspect(sort(whs_rules, by ='lift'))

```


We start by looking at the data region by region to find any patterns in the 
associations.
```{r , comment= TRUE}
wholesale_Lisbon = whs_convertion %>% filter(RLisbon == 1)
wholesale_Oporto = whs_convertion %>% filter(ROporto == 1)
wholesale_Other = whs_convertion %>% filter(ROther == 1)

# Region LisbonH support, confidence, and lift (still using overall data's high,
# medium, and low levels):
LisbonH = wholesale_Lisbon %>% select(CHoreca, CRetail, FreshH, MilkH, GroceryH,
                                      FrozenH, DetergentsH, DelicatessenH)
LisbonH = LisbonH %>% mutate(across(,as.logical))
LisbonH_rules = apriori(LisbonH, parameter = list(supp=0.15,conf=0.7))
inspect(sort(LisbonH_rules, by ='lift'))
```

Across all regions and channels, the high spending clients who buy products from
the Milk and Detergents category are 340% more likely to purchase products from
the Grocery category than pure chance. 

Across all region, high spending clients from Retail who buy products from
Grocery are 3 times more likely to purchase products from
the Detergents category and vice versa than pure chance. 

High spending clients who buy products from Grocery and Detergents are 3 times
more likely to be from the Retail channel.

```{r , comment= TRUE}
# Region OportoH support, confidence, and lift (still using overall data's high,
#medium, and low levels):
OportoH = wholesale_Oporto %>% select(CHoreca, CRetail, FreshH, MilkH, GroceryH,
                                      FrozenH, DetergentsH, DelicatessenH)
OportoH = OportoH %>% mutate(across(,as.logical))
OportoH_rules = apriori(OportoH, 
                        parameter = list(supp=0.1,conf=0.7))
inspect(sort(OportoH_rules, by ='lift'))

```


```{r , comment= TRUE}
# Region OtherH support,confidence, and lift (still using overall data's high, 
# medium, and low levels):
OtherH = wholesale_Other %>% select(CHoreca, CRetail, FreshH, MilkH, GroceryH, 
                                    FrozenH, DetergentsH, DelicatessenH)
OtherH = OtherH %>% mutate(across(,as.logical))
OtherH_rules = apriori(OtherH, 
                       parameter = list(supp=0.1, conf=0.7, minlen = 2))
inspect(sort(OtherH_rules, by ='lift'))
```
By looking at the lift values for the different regions, we found that across 
all regions and channels, the high spending clients from the Retail channel who
buy products from the Milk and Detergents category are 3 times more likely to 
purchase products from the Grocery category than pure chance. 


```{r }
# Look into Channel regardless of regions:
whs_Horeca =  whs_convertion %>% filter(CHoreca == 1)
whs_Retail = whs_convertion %>% filter(CRetail == 1)


# Horeca support,confidence, and lift:
HorecaH = whs_Horeca %>% select(FreshH, MilkH, GroceryH, FrozenH, DetergentsH, 
                                DelicatessenH) %>% 
  mutate(across(,as.logical))
HorecaH_rules = apriori(HorecaH, 
                        parameter = list(supp=0.1, conf=0.4,minlen = 2))
inspect(sort(HorecaH_rules, by ='lift'))

# Retail support, confidence, and lift:
RetailH = whs_Retail %>% select(FreshH, MilkH, GroceryH, FrozenH, DetergentsH, 
                                DelicatessenH) %>% 
  mutate(across(,as.logical))
RetailH_rules = apriori(RetailH, 
                        parameter = list(supp=0.4, conf=0.8, minlen = 2))
inspect(sort(RetailH_rules, by ='lift'))
```


```{r }
# Look into Channel regardless of regions:

whs_Horeca =  whs_convertion %>% filter(CHoreca == 1)
whs_Retail = whs_convertion %>% filter(CRetail == 1)

```


```{r }
HorecaH = whs_Horeca %>% select(FreshH, MilkH, GroceryH, FrozenH, DetergentsH, DelicatessenH) %>% 
  mutate(across(,as.logical))

HorecaH_rules = apriori(HorecaH, 
                        parameter = list(supp=0.1, conf=0.4,minlen = 2))

inspect(sort(HorecaH_rules, by ='lift'))
```

For the high spenders in the Horeca channel, clients who buy products from the
Delicatessen category are 90% more likely to purchase products from the Frozen
category than pure chance. 

```{r }
RetailH = whs_Retail %>% select(FreshH, MilkH, GroceryH, FrozenH, DetergentsH,
                                DelicatessenH) %>% 
  mutate(across(,as.logical))

RetailH_rules = apriori(RetailH, 
                        parameter = list(supp=0.4, conf=0.8, minlen = 2))

inspect(sort(RetailH_rules, by ='lift'))

```
For the high spenders in the Retail channel, clients who buy products from the
Milk and Detergents category are 58% more likely to purchase products from the
Grocery category than pure chance.




#### 4.2 Clustering

Clustering technique helps find meaningful client segments. 

##### 4.2.1 Outlier Removal:
Looking at the data, we noticed that there are a few values in each column that 
are much greater than the rest of the values in each column. So, we plotted 
histograms for the upper end of values in each category(from the mid point of 
each category to the maximum value of that category) to see how the data is 
distributed and to set an appropriate upper threshold as clustering is sensitive
to outliers.
```{r }
# Converting the data to a key value format and plotting the histograms for the
# 6 categories
cust_data_bin = wholesale %>% filter(Fresh>(max(Fresh)/2) | Milk>(max(Milk)/2)|
                                     Grocery>(max(Grocery)/2)|
                                     Frozen>(max(Frozen)/2)|
                                     Detergents_Paper>(max(Detergents_Paper)/2)|
                                     Delicatessen>(max(Delicatessen)/2)) %>% 
  select(Fresh,Milk,Grocery,Frozen,Detergents_Paper,Delicatessen)

ggplot(gather(cust_data_bin), aes(value)) + 
  geom_histogram(bins = 15) + 
  facet_wrap(~key, scales = 'free_x')

``` 


##### Cleaning the data by removing the outliers and adding the names for regions and channels:
```{r}
cust_ol_filt = wholesale %>% 
  filter(Fresh<65000,Milk<50000,Grocery<68000,Frozen<30000,
         Detergents_Paper<30000,Delicatessen<20000) %>% 
  mutate(region_name = case_when(Region==1 ~ 'Lisbon',Region==2 ~ 'Oporto',
                                 Region==3 ~ 'Other'),
         channel_name = case_when(Channel==1 ~ 'Horeca',
                                  Channel==2 ~ 'Retail'))
```

##### 4.2.2 Data normalization
##### Normalizing the data and plotting the SSE curve for finding the optimum number of clusters:
```{r}
# Using Min-max normalization
normalize = function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

cust_normalized = cust_ol_filt %>% mutate_at(c(3:8), normalize)
distance_matrix = dist(cust_normalized[,3:8], method = 'euclidean')

sse_curve = c()
for (n in 1:10) {
  cust_kcluster = kmeans(cust_normalized[,3:8], n)
  sse = cust_kcluster$tot.withinss
  sse_curve[n] = sse}
plot(1:10, sse_curve, type = 'b')
```

We are choosing 5 clusters as it has a low SSE value and any cluster size above 
5 gives us clusters with very sparse items.

```{r}
# Performing clustering with 5 clusters
cust_clust = kmeans(cust_normalized[,3:8], centers = 5)
sil_co = silhouette(cust_clust$cluster, dist = distance_matrix)
cust_clust$centers
summary(sil_co)
```


##### 4.2.3 Denormalization of values:
Since all the values are from 0 to 1, we are denormalizing them to their actual
amounts.
```{r}
cust_clust_df=as.data.frame.matrix(cust_clust$centers)

# Creating a function for denormalization
denormalize <- function(x,minval,maxval) {
    return(x*(maxval-minval) + minval)}

cust_clust_denorm=data.frame(matrix(ncol = 0, nrow = 5))

# Passing each column and denormalizing the values
for (i in 1:ncol(cust_clust_df)) {
  max_val = max(cust_ol_filt[i+2])
  min_val = min(cust_ol_filt[i+2])
  cust_clust_denorm[i] = denormalize(cust_clust_df[i], min_val, max_val)
}
cust_clust_denorm
```


##### 4.2.4 Cluster Results:
We are able to see 5 clusters with the following attributes:\
Cluster 1: High spenders across most categories\
Cluster 2: High Fresh and Frozen spenders.\
Cluster 3: Low spenders across all categories.\
Cluster 4: High Fresh spenders.\
Cluster 5: High Grocery spenders.



Based on the results of the clustering analysis, we have grouped clients into 
the following five groups:\

● High Spenders: The 19 clients from this group spend the most across all the categories.
In addition, all these consumers belong to the Retail channel.

● High Fresh and Frozen Spenders: The 44 clients from this group spend more on
products from the ‘Fresh’ and ‘Frozen’ category than the other categories and 93% of
them are from the Horeca channel.

● Low Spenders: The 210 clients in this group spend less than other clients for all product
categories. Among the low spenders, 89% of this group are from the Horeca channel.

● High Fresh Spenders: The 62 clients from this group purchase the most amount within
the ‘Fresh’ category amongst all clients compared to other categories. Also, 82% of this
group of clients are from the Horeca channel.

● High Grocery Spenders: The 95 clients from this group purchase more products from
the ‘Grocery’ category compared to the other categories. 85% of them are from the Retail
channel.



### 5. Key Recommendations

● When a client purchases products from one of the ‘Milk’, ‘Detergents’, and ‘Grocery’
categories, recommend products from the other two categories which can lead to
increased sales, especially for clients from the Retail channel. For the Horeca channel, it
is beneficial to recommend products from the ‘Frozen’ categories when clients purchase
products from ‘Delicatessen’. It is imperative to make sure there is no gap in the
inventory level of these categories.


● ‘Fresh’, ‘Frozen ', and ‘Grocery’ are the highest selling categories, therefore a minimum
inventory threshold should be set to ensure these categories can be replenished in time.
Furthermore, there is no need for discounts on these categories due to the high sales.
Instead, provide discounts on ‘Delicatessen’, ‘Milk’, and ‘Detergents’ as incentive to
increase sales.


